Type Error Telemetry at Scale

### 2023-09-22 programming23/paper.tex
comments and a bit of fixing

@@

\Cref{f:records-per-hour} provides a time-ordered distribution of the data.
Each vertical bar counts the number of records generated per hour according to
a client-side timestamp.
The bars are labeled with a California time zone because that is the locale of
the Roblox servers, which parsed the timestamps.
Most buckets contain roughly one thousand records, but some have very few
and a handful of others exceed three thousand.
%% \FILL{} investigate one tower, what's happening?? all one session?
There is a sharp drop in early April when the experiment was finished
and the telemetry probe was removed (though there is a long tail as
developers uopdated Roblox Studio).
We do not know which timezone the records originated in, but since the counts
tend to peak midday California time it seems that the majority
of Roblox developers are following a Western Hemisphere schedule.
The counts for weekends (shaded regions) are often the tallest,
as Roblox has a significant school-aged creator community.

@@

A close inspection of the data revealed a few anomalies.
For the most part, we omit the strange data from futher analysis.

First, some records have odd timestamps according to the server's records.
Every telemetry record comes with two timestamps: a client-side timestamp
from when Roblox Studio created it and a server-side timestamp from when
the server posted it to its database.
The server timestamps are little use for our analysis because they bunch
together; it is common for several records to have identical server
timestamps.
However, server timestamps are useful for auditing because they should
be relatively close in value to the client timestamps.
For a small number of records (<100? \FILL{}), though, the client is off by a large
margin: up to one week behind and up to one day ahead of the server.
We attribute the large delays to issues when sending telemetry (lost internet
connection, etc.) and smaller offsets to incorrect time settings on the client
machine.
In any event, we do not filter these records from further analysis.

@@

The first step toward understanding the data is to have a sense of what
is in it.
Three important characteristics are the size of the codebases in the data,
the length of the sessions we captured, and the number of type analysis
errors.
We discuss these in turn.
Codebase size and session length follow power-law distributions, with
many small items and a few very large items.
The number of type analysis errors is somewhat low~(explained in the next
section), but sufficient to proceed with futher analysis.

@@

  \caption{Size of analyzed code: number of files, number of lines , and lines in edit range}
  \label{f:codebase-size}
\end{figure}

\Cref{f:codebase-size} summarizes the size of projects in the dataset.
The top half of the figure reports the mean, standard deviation, median,
and \percentile{99} for the number of files, lines, and lines in edit range.
The bottom half shows zoomed-in distributions of the file, line, and edit range
numbers.
For example, the x-axis of the first plot ranges from 0 to 1,000 files and
the y-axis counts up to 1,000 telemetry records.

There is a huge amount of variation across projects.
The largest ones have over 50,000 files, and
there are many small ones with 1 file and/or 1 line of code.
Unsurprisingly, the numbers come with large standard deviations.
The median values are more reasonable, with roughly 3,000 lines of code
and 1,000 lines in edit ranges, but still have a large number of files: over 7,000.
One source of variation is that some developers use Luau scripts as a way
to store data, writing a module that exports one huge array, typically
defined on a single line. Such data storage scripts provide a challenge for
script tooling.

@@

\Cref{f:session-size} outlines the size of the Roblox Studio sessions in the
data.
There are two useful notions of size: real time in seconds (between the first
and last record) and the total number of records.
The top of~\cref{f:session-size} reports the mean, standard deviation,
median, and \percentile{99}; the bottom shows size distributions, zoomed in
to focus on at most 1,000 units (x-axis) and 1,000 sessions (y-axis).

@@

\begin{figure}[t]
  \begin{tabular}[t]{ll} \\
    \begin{tabular}[t]{l@{~~}r@{~}l}
      595,137 & type errors \\
      289,698 & in current module & [\pct{48.68}] \\
       30,924 & in edit range & [\pct{5.20}]
    \end{tabular}
    \begin{tabular}[t]{l@{~~}r@{~}l}
      72,235,735 & {forced strict errors} \\
      37,027,281 & in current module & [\pct{51.26}] \\
       1,111,178 & in edit range & [\pct{1.54}]
    \end{tabular}
  \end{tabular}
  \caption{Type errors and forced strict errors across all telemetry records}
  \label{f:count-analysis-errors}
\end{figure}

@@

\Cref{f:count-analysis-errors} lists the number of type analysis errors
and breaks them down by location in the codebase.
In total, there are nearly 600K type errors.
These appeared in modules that opted in to typechecking with either the
\mnonstrict{} or \mstrict{} mode.
The number of forced strict errors is much larger, 72M, because this analysis
ran on every module using strict type.
(Furthermore, forced strict runs silently, so creators have no incentive to
fix these errors.)
Approximately half the analysis errors occurred in the current module,
and a small fraction overlapped with the current edit range~(\pct{5} for
type errors, \pct{1} for forced strict).

@@

  \begin{subfigure}[t]{\columnwidth}
    \begin{tabular}[t]{ll} \\
      \begin{tabular}[t]{l@{~~}r@{~}l}
        313,509 & \mnocheck{}   & [\pct{90.19}] \\
         32,902 & \mnonstrict{} & [\pct{ 9.47}] \\
            545 & \mstrict{}    & [\pct{ 0.16}] \\
            642 & mixed-mode    & [\pct{ 0.18}]
      \end{tabular}
      \begin{tabular}[t]{l@{~~}ll}
        \zerowidth{Among the mixed-mode sessions:} \\
        & 341 & contain a mode upgrade \\
        & 320 & contain a mode downgrade \\
        & 512 & contain modules with different modes
      \end{tabular}
    \end{tabular}
    \caption{Analysis mode(s) for the 347,598 sessions}
    \label{f:total-sessions}
  \end{subfigure}

@@

%% mostly pure, rare mixes
Most sessions (\pct{99.82}) stick to a single analysis mode.
These sessions never change the mode in the current module and never switch
focus to a module with a different analysis mode.
Among the mixed-mode sessions, most (\pct{80}) switch to modules with
a different mode, about half contain at least one edit that upgrades the
current mode, and about half contain a downgrade.
There are 619 upgrades in total and 583 downgrades.
\FILL{} Ben check how often the downgrades follow the upgrades.

@@

A possible explanation for the low adoption of \mnonstrict{} and
\mstrict{} mode is that upgrading to these modes leads to a large
number of type errors.
Creators might get discouraged by a high error count and revert to
\mnocheck{} mode.

@@

Similarly, mode downgrades have only a small negative effect on the
number of errors (mean -0.3, stddev 4, median 0, max -48).
%% curious outlier: downgrade added 30 errors!
Creators apparently tend switch modes only when the code is already
well typed, or close to it.

@@

Equipped with an overall picture of the dataset and the skewed usage
of \mnocheck{}, \mnonstrict{}, and \mstrict{} modes, we are ready to explore
the type error data.
Our main focus is on the intersection between type errors
and edits to the programs.
When a type error highlights part of the code and the next edit range includes
that code, we want to know whether the error survives the edit.
Similarly, we want to know when an edit introduces a type error.

@@

%% \paragraph{Most Type Errors are Typos}
\begin{itemize}
  \item
    One striking aspect of~\cref{t:type-error-survival} is the large number
    of \code{SyntaxError}s and the related errors \code{UnknownSymbol} and
    \code{UnknownProperty}.
    Most type errors in the data are likely due to typos.
    \Cref{s:type-error-count} discusses this point in detail.
    For now, we remark that \code{SyntaxError} is the only possible error
    in \mnocheck{} mode (by design) and observe that these errors are
    usually in the Drop column and seldom in the Keep column.

@@

To deal with pathologies such as the worst-case time for ML-style type
inference~\cite{m-popl-1990,ktu-caap-1990}, the typechecker
has internal limits that restrict the problems it will attempt to solve.
Hitting a limit triggers one of the following errors:
\code{CodeTooComplex},
\code{NormalizationTooComplex}, or
\code{UnificationTooComplex}.
A casual user should never see these errors.

>>>>

\usepackage{todonotes}

@@

\Cref{f:records-per-hour} provides a time-ordered distribution of the data.
Each vertical bar counts the number of records generated per hour according to
a client-side timestamp.\todo{Do we need all this detail?}
The bars are labeled with a California time zone because that is the locale of
the Roblox servers, which parsed the timestamps.
Most buckets contain roughly one thousand records, but some have very few
and a handful of others exceed three thousand.
%% \FILL{} investigate one tower, what's happening?? all one session?
There is a sharp drop in early April when the experiment was finished
and the telemetry probe was removed (though there is a long tail as
developers uopdated Roblox Studio).
We do not know which timezone the records originated in, but since the counts
tend to peak midday California time it seems that the majority
of Roblox developers are following a Western Hemisphere schedule.
The counts for weekends (shaded regions) are often the tallest,
as Roblox has a significant school-aged creator community.

@@

A close inspection of the data revealed a few anomalies.
For the most part, we omit the strange data from futher
analysis.\todo{I'm inclined to rewrite this to be much more concise.}

First, some records have odd timestamps according to the server's records.
Every telemetry record comes with two timestamps: a client-side timestamp
from when Roblox Studio created it and a server-side timestamp from when
the server posted it to its database.
The server timestamps are little use for our analysis because they bunch
together; it is common for several records to have identical server
timestamps.
However, server timestamps are useful for auditing because they should
be relatively close in value to the client timestamps.
For a small number of records (<100? \FILL{}), though, the client is off by a large
margin: up to one week behind and up to one day ahead of the server.
We attribute the large delays to issues when sending telemetry (lost internet
connection, etc.) and smaller offsets to incorrect time settings on the client
machine.
In any event, we do not filter these records from further
analysis.\todo{To be clear, you do NOT filter? Because the section is
  about what you DO filter.}

@@

Three important characteristics of the recorded data are the size of the codebases,
the length of the sessions, and the number of type
errors.
We discuss these in turn.
Codebase size and session length follow power-law
distributions,\todo{Please confirm w/ a statistical test. See Clauset
  et al.: arXiv:0706.1062.} with
many small items and a few very large items.
The number of type analysis errors is somewhat low~(explained in the next
section), but sufficient to proceed with futher analysis.

@@

  \caption{Size of analyzed code: number of files, number of lines, and lines in edit range}
  \label{f:codebase-size}
\end{figure}

\Cref{f:codebase-size} summarizes the size of projects in the dataset.
The top half of the figure reports the mean, standard deviation, median,
and \percentile{99} for the number of files,\todo{There are
  1000s of files?!? More files than lines? Makes no sense.} lines, and lines in edit range.
The bottom half shows zoomed-in distributions of the file, line, and edit range
numbers.
For example, the x-axis of the first plot ranges from 0 to 1,000 files and
the y-axis counts up to 1,000 telemetry records.

There is a huge amount of variation across projects.
The largest ones have over 50,000 files, and
there are many small ones with 1 file and/or 1 line of code.
Unsurprisingly, the numbers come with large standard
deviations.\todo{Which is not shown!}
The median values are more reasonable, with roughly 3,000 lines of code
and 1,000 lines in edit ranges, but still have a large number of files: over 7,000.
One source of variation is that some developers use Luau scripts as a way
to store data, writing a module that exports one huge array, typically
defined on a single line. Such data storage scripts provide a challenge for
script tooling.\todo{No idea what's said here.}

@@

\Cref{f:session-size} outlines the size of the Roblox Studio sessions in the
data.
There are two useful notions of size: real time in seconds (between the first
and last record) and the total number of records.
The top of~\cref{f:session-size} reports the mean, standard deviation,
median, and \percentile{99}; the bottom shows size distributions, zoomed in
to focus on at most 1,000 units (x-axis) and 1,000 sessions
(y-axis).\todo{Again, confused.}

@@

\begin{figure}[t]
  \begin{tabular}[t]{ll} \\
    \begin{tabular}[t]{r@{~~}r@{~}l}
      595,137 & type errors \\
      289,698 & in current module & [\pct{48.68}] \\
       30,924 & in edit range & [\pct{5.20}]
    \end{tabular}
    \begin{tabular}[t]{r@{~~}r@{~}l}
      72,235,735 & {forced strict errors} \\
      37,027,281 & in current module & [\pct{51.26}] \\
       1,111,178 & in edit range & [\pct{1.54}]
    \end{tabular}
  \end{tabular}
  \caption{Type errors and forced strict errors across all telemetry records}
  \label{f:count-analysis-errors}
\end{figure}

@@

\Cref{f:count-analysis-errors} lists the number of type analysis errors
and breaks them down by location in the codebase.
In total, there are nearly 600K type errors.
These appeared in modules that opted in to typechecking with either the
\mnonstrict{} or \mstrict{} mode.
The number of forced strict errors is much larger, 72M, because this analysis
ran on every module using strict type.
(Furthermore, forced strict runs silently, so creators have no
awareness of or incentive to
fix these errors.)
Approximately half the analysis errors occurred in the current module,
and a small fraction overlapped with the current edit range~(\pct{5} for
type errors, \pct{1} for forced strict).

@@

  \begin{subfigure}[t]{\columnwidth}
    \begin{tabular}[t]{ll} \\
      \begin{tabular}[t]{r@{~~}r@{~}r}
        313,509 & \mnocheck{}   & [\pct{90.19}] \\
         32,902 & \mnonstrict{} & [\pct{ 9.47}] \\
            545 & \mstrict{}    & [\pct{ 0.16}] \\
            642 & mixed-mode    & [\pct{ 0.18}]
      \end{tabular}
      \begin{tabular}[t]{l@{~~}ll}
        \zerowidth{Among the mixed-mode sessions:} \\
        & 341 & contain a mode upgrade \\
        & 320 & contain a mode downgrade \\
        & 512 & contain modules with different modes
      \end{tabular}
    \end{tabular}
    \caption{Analysis mode(s) for the 347,598 sessions}
    \label{f:total-sessions}
  \end{subfigure}

@@

%% mostly pure, rare mixes
Most sessions (\pct{99.82}) stick to a single analysis mode.
These sessions never change the mode in the current module and never switch
focus to a module with a different analysis mode.
Among the mixed-mode sessions, most (\pct{80}) switch to modules with
a different mode, about half contain at least one edit that upgrades
the typing strictness in the
current mode, and about half contain a downgrade.
There are 619 upgrades in total and 583 downgrades.
\FILL{} Ben check how often the downgrades follow the upgrades.

@@

A possible explanation for the low adoption of \mnonstrict{} and
\mstrict{} mode is that upgrading to these modes leads to a large
number of type errors.
Creators might get discouraged or overwhelmed by a high error count and revert to
\mnocheck{} mode.

@@

Similarly, mode downgrades have only a small negative effect on the
number of errors (mean -0.3, stddev 4, median 0, max -48).
%% curious outlier: downgrade added 30 errors!
Creators apparently tend switch modes only when the code is already
well typed, or close to it.\todo{What incentive is there to do so?}

@@

Equipped with an overall picture of the dataset and the skewed usage
of \mnocheck{}, \mnonstrict{}, and \mstrict{} modes, we are ready to explore
the type error data.
Our main focus is on the intersection between type errors
and edits to the programs.\todo{The next two sentences are in some
  sense our main RQs, right? I would call them out more explicitly and
  w/ more justification.}
When a type error highlights part of the code and the next edit range includes
that code, we want to know whether the error survives the edit.
Similarly, we want to know when an edit introduces a type error.

@@

%% \paragraph{Most Type Errors are Typos}
\begin{itemize}
  \item
    One striking aspect of~\cref{t:type-error-survival} is the large number
    of \code{SyntaxError}s and the related errors \code{UnknownSymbol} and
    \code{UnknownProperty}.
    Most type errors in the data are likely due to typos.
    \Cref{s:type-error-count} discusses this point in detail.
    For now, we remark that \code{SyntaxError} is the only possible error
    in \mnocheck{} mode (by design) and observe that these errors are
    usually in the Drop column and seldom in the Keep column.\todo{So?}

@@

To deal with pathologies such as the worst-case time for ML-style type
inference~\cite{m-popl-1990,ktu-caap-1990}, the typechecker
has internal limits that restrict the problems it will attempt to solve.
Hitting a limit triggers one of the following errors:
\code{CodeTooComplex},
\code{NormalizationTooComplex}, or
\code{UnificationTooComplex}.
A casual user should never see these errors.\todo{IDEALLY never see?
  They CAN see them, right?}
